Safe Memory Regions for Big Data Processing

### 2016-06-13 paper/type-inference.tex
type inference

@@

@@

\subsection{Modular Constraint Solving}
\label{sec:fb-constraintsolving}

Our constraint generation algorithm traverses the entire program,
performing elaboration and collecting constraints, which are
subsequently solved en masse. The motivation behind the whole-program
approach to constraint generation is twofold: it simplifies
elaboration functions and makes presentation easier, and second, it
naturally generalizes to mutual recursion. Nonetheless, we do not
intend our type inference to be a whole-program analysis for (a). it
preempts opportunities for separate compilation and dynamic linking,
and (b). it is expensive and an overkill in most practical cases. We
therefore reclaim the compositionality of type inference by solving
the constraints in a compositional fashion. In more practical terms
this means that our constraint solving algorithm visits and solves
every constraint (or, every set of mutually dependent constraints)
only once. It composes computed solutions to solve other constraints
that depend on the solved constraints. Importantly, the failure to
solve a dependent constraint does not result in backtracking. We now
describe our compositional constraint solving algorithm in detail. To
simplify the presentation of our algorithm, we assume that are no
mutually recursive definitions in the source program. Recursive
definitions are nonetheless allowed.

\paragraph{Terminology} In a validity constraint, a predicate variable
occuring on the left side of the turnstile is said to occur
negatively, or with \emph{negative polarity}. In contrast, a predicate
variable occuring on the right side is said to occur positively, or
with \emph{positive polarity}. A validity constraint
\emph{constrains} the set of predicate variables that occur
negatively in the constraint, while it \emph{uses} the set of
predicate variables that occur positively. A constraint is said to be
\emph{recursive} if it constrains and uses a predicate variable.
% A pair of validity
% constraints are said to be \emph{mutually dependent} if there exists a
% pair ($\varphi_1$, $\varphi_2$) of predicate variables that occur
% with opposite polarities in both the constraints. In such case, the
% pair ($\varphi_1$,$\varphi_2$) of predicate variables are also said to
% be mutually dependent. We call the transitive closure of mutual
% dependency relation as \emph{transitive dependency} relation. Like
% mutual dependency, transitive dependency is also extended to predicate
% variables.

Given a set of validity constraints, we first build a dependency graph
($G_c$) with constraints as nodes, and dependencies between them
captured as edges. There exists an edge from a constraint $c_2$ to a
constraint $c_1$ in the graph (i.e., $c_2$ \emph{depends on} $c_1$) if
$c_1$ constrains a predicate variable that $c_2$ uses.

The above condition intuitively corresponds to a case, where
expression or type, whose region elaboration is constrained by $c_2$
refers to a method or a class, whose unknown precondition is
constrained by $c_1$. The common predicate variable ($\varphi$)
represents the unknown precondition in this case. The dependency from
$c_2$ to $c_1$ means that $c_1$ must be solved to compute $\varphi$
before $c_2$ is solved, thus enforcing the rule that the precondition
of a method must not depend on its calling context. 
% Note that the first condition results in self-loops over recursive
% constraints in the dependency graph.  Since mutually recursive
% definitions refer each other, the first condition also results in
% bidirectional dependencies between mutually recursive constraints
% (and self-loops over recursive constraints).

Next, we convert the dependency graph over constraints into a
dependency DAG ($G_C$) over sets of constraints, where each set
represents a strongly connected component in the dependency graph.

%% \paragraph{Example} The dependency DAG ($G_C$) over validity
%% constraints from the \C{Pair} example (\S~\ref{sec:constraints}) is
%% shown below: %in Fig.~\ref{fig:pair-deps}. 
%% \vspace*{-0.08in}
%% \begin{figure}[H]
%% \includegraphics[scale=0.6]{DepDAG.png}
%% \end{figure}
%% \vspace*{-0.08in}
%% Each node (labeled $C_i$) is a set of constraints that belong to a
%% strongly connected component in the dependency graph ($G_c$), hence
%% are mutually dependent. All dependencies, except the self-dependency
%% on $\mathbf{c_{10}}$, are type-2 dependencies.

A dependency DAG makes the dependencies between constraints explicit.
Constraints in each set are mutually dependent, and need to be solved
simultaneously, whereas constraints in different sets can be solved as
per any valid topological ordering of the graph's transpose.
Accordingly, we obtain a topological ordering of nodes in the graph
$G_{{C}}^{T}$ ($G_C$'s transpose), and solve the sets of constraints
in that order. The solutions obtained after solving a constraint set
are applied to the constraints in subsequent sets before attempting to
solve them.  Consequently, when the turn of a constraint set ($C$)
arrives during the constraint solving process, it satisfies certain
properties:
\begin{itemize}
\item There exists only one predicate variable ($\varphi$) that is either
constrained or used by the constraints in the set ($C$). The variable is
called set's \emph{subject}. This property follows from (a). the fact
that all the dependency constraints have already been solved (and
solutions applied), and (b). the assumption that there are no mutually
recursive definitions. 
\item All the constraints that constrain the set's subject are present
in the set. This follows from our definition of the dependency relation.
\end{itemize}

%% For the DAG in figure above, we consider the topological order $[C_1,
%% C_2, C_3]$ of its transpose, and solve the sets of constraints in that
%% order.

\subsection{Expression Elaboration}

Function $\elabExpr$, shown in Fig.~\ref{fig:fb-elabexpr}, performs
this elaboration for a subset of expressions in $\absof{\FB}$, whose
corresponding $\FB$ expressions have been ascribed static semantics in
Fig.~\ref{fig:fb-staticsem}. $\elabExpr$ is defined under the same
context as the expression typing judgment in
Fig.~\ref{fig:fb-staticsem} with symbols $\A$,$\ralloc$, and $\env$
retaining their meaning. The function traverses expressions in a
syntax-directed manner of a type checker, introducing fresh region
type templates for unknown region types, while generating constraints
over region and predicate variables. The precise nature of generated
constraints is explained in \S\ref{sec:fb-constraintsem}, but in
summary, they capture the relationships between the type templates of
various subexpressions and their well-formedness. Note that
$\elabExpr$ returns the region type template of the subexpression,
which is used to generate constraints for the expression. Functions
$\typeOk$ and $\subtypeOk$ (definitions not shown) used by $\elabExpr$
implement type well-formedness and subtype judgments from
Fig.~\ref{fig:fb-staticsem}, respectively.

\subsection{Method and Class Elaboration}

Function $\elabMeth$ elaborates a method definition of class $B$. It
calls $\elabExpr$ with the context $\A$, its allocation context
parameter ($\rhoallocm$), and a type environment ($\env$) that
contains region type bindings for all the arguments of the method,
including the implicit $\C{this}$ argument. The region type template
returned by $\elabExpr$ for the method body is checked against its
expected type (derived from the type template of the method)
generating more constraints. The function then returns the elaborated
method definition and the set of constraints.

>>>>

In the elaboration phase, we can process a class \C{C} only after any class
\C{B} that \C{C} depends on has been processed: class \C{C} depends on
class \C{B} if \C{B} is either \C{C}'s base class or the type of any field
of \C{C} depends on \C{B}. In effect, this means that any collection of
mutually recursive classes must be processed together. Non-recursive
dependences can be handled in a compositional fashion: if class \C{C}
depends on \C{B} non-recursively, then the elaboration can be done for
\C{B} first, and then \C{C} can be processed.

The same idea applies to the constraint-solving phase as well.
Given a set of constraints, we say that a predicate variable $\varphi_1$
\emph{directly-depends} on another predicate variable $\varphi_2$ if the set of
constraints includes a constraint $\isvalid{\varphi_1 \conj \phictxt}{F(\varphi_2)}$.
We say that $\varphi_1$ \emph{depends} on $\varphi_2$ if $\varphi_1$ transitively
depends on $\varphi_2$.
The constraint solver needs to process any collection of mutually dependent
predicate variables together.
In effect, this requires the type inference to process any collection of
mutually recursive methods together.
However, methods that are not mutually recursive can be processed separately.

@@

\subsection{Elaboration and Constraint Generation}

We now present an integrated algorithm that processes all the methods
in a given program, elaborating them (by introducing region variables
to stand for actual region parameters) and simultaneously generating
the set of constraints that must be solved.

Function $\elabExpr$, shown in Fig.~\ref{fig:fb-elabexpr}, performs
this elaboration for a subset of expressions in $\absof{\FB}$, whose
corresponding $\FB$ expressions have been ascribed static semantics in
Fig.~\ref{fig:fb-staticsem}. $\elabExpr$ is defined under the same
context as the expression typing judgment in
Fig.~\ref{fig:fb-staticsem} with symbols $\A$,$\ralloc$, and $\env$
retaining their meaning. The function traverses expressions in a
syntax-directed manner of a type checker, introducing fresh region
type templates for unknown region types, while generating constraints
over region and predicate variables.
Note that $\elabExpr$ returns the type of the subexpression,
which is used to generate constraints for the expression. Functions
$\typeOk$ and $\subtypeOk$ (definitions not shown) used by $\elabExpr$
implement type well-formedness and subtype judgments from
Fig.~\ref{fig:fb-staticsem}, respectively.

Function $\elabMeth$ elaborates a method definition of class $B$. It
calls $\elabExpr$ with the context $\A$, its allocation context
parameter ($\rhoallocm$), and a type environment ($\env$) that
contains region type bindings for all the arguments of the method,
including the implicit $\C{this}$ argument. The type 
returned by $\elabExpr$ for the method body is checked against its
expected type (derived from the type of the method)
generating more constraints. The function then returns the elaborated
method definition and the set of constraints.

@@
